---
title: "Coding Concepts"
output:
  pdf_document: default
  ioslides_presentation:
    highlight: pygments
  slidy_presentation:
    highlight: pygments
  html_document: default
  beamer_presentation:
    highlight: pygments
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
#  Implementing Functions in R 

Format for a basic function in R

<span style="color: orange;"> #' documentation that describes inputs, outputs and what the function does

<span style="color: orange;"> FUNCTION NAME = function(inputs, parameters) {

<span style="color: orange;">body of the function (manipulation of inputs)

<span style="color: orange;">return }</span>

In R, inputs and parameters are treated the same; but it is useful to think about them separately in designing the model - collectively they are sometimes referred to as arguments

#  Best practices for model (software) development 

## Common problems

* Unreadable code (hard to understand, easy to forget how it works, hard to find errors, hard to expand)

* Overly complex, disorganized code (hard to find errors; hard to modify-expand)

* Insufficient testing (both during development and after)

* Not tracking code changes (multiple versions, which is correct?)



#  Best practices for model (software) development 

Basic Solution

Structured practices that ensures
clear, readable code 

* meaningful variables names 

* comments to explain the workflow

* documentation to explain functionality 

* modularity (organized “independent” building blocks)

*  testing frequently as you develop each module and after

* code evolution is documented (version control)




#  Best practices for model (software) development 

" *Let us change our traditional attitude to the construction of programs: Instead of imagining that our main task is to instruct a computer what to do, let us concentrate rather on explaining to humans what we want the computer to do. -- Donald E. Knuth, Literate Programming, 1984* "

Developing readable (by PEOPLE) code and documenting what you are doing is essential

*“When was the last time you spent a pleasant evening in a comfortable chair, reading a good program?”— Bentley (1986)*



#  Documentation! 

Makes it easier for you to fix issues, refine and easier for others to use

For functions (model) you should always include the following in your code

* The goal of the model

 * Model inputs (description, name and units)

* Model output (description, name and units)

* Where the model (or submodel) comes from (source papers)

In line explanation for key steps

Always use meaningful names for functions, variables, inputs and outputs

Always put your function in its own file - with just the documentation





#   Building Models 

Another example: Okuns Law (conceptual, abstract model) - try implementing this as a function

Graph of US quarterly data (not annualized) from 1947 through 2002 estimates a form of the difference version of 

Okun's law:

**%Change GNP = 3.2 - 1.8(Change Unemployment Rate)** 

R\^2 of .463. 

Differences from other results are partly due to the use of quarterly data
http://en.wikipedia.org/wiki/Okun%27s_law

Write the function in R!

what is change in GDP for unemployment rate change of 3%

what would the change in GDP be for 3% change in unemployment rate if the slope was 10% steeper


```{r, okum example}

# from Okum's Law - change in GDP is a linear function of change in unemployment rate
# note for teaching I'm coding this inline but a good practice is to code the function in its own R file
compute_GDP_change = function( unemploy_change, slope=-1.8, intercept=3.2) {
  delta_GDP = intercept + slope*unemploy_change
  return(delta_GDP)
}

# now lets run the function to see what change in GDP is for 3% change in unemployment rate

compute_GDP_change(3)

# what if unemployment was even more sensitive
compute_GDP_change(3, slope = -10)

```




---

#  Building Models:Documentation 

There is also a standard format for documentation that can be read by automatic programs (roxygen2) - an R package that generate “standard” R documentation - manual or help pages

These automated approaches for building documentation (like roxygen2) and meta data (descriptions of data sets) are increasingly common - so you should get into the practice of being structured in your approach to documentation

We will use the conventions that work with roxygen2 - and then use this program to generate formal R documentation. Roxygen is similar to Doxygen which is used for C code…so its a widely used format

---

# Lists (Short Review) - 

For returning more than one item from your function!

* Lists are the most “informal” data structures in R

* List are really useful for keeping track of and organizing groups of things that are not all the same

* A list could be a table where number of rows is different for each column

* A list can have numeric, character, factors all mixed together

*  List are often used for returning more complex information from function (e.g. lm)

# Examples 

Here are some examples of how to create and manipulate lists

```{r lists}
# simple list
sale =  list(number=c(2,4,9), quality="high", what="apple", cost=4)
sale
sale$cost
sale$number
sale[4]

```

# Lists and Functions

Lists are used to return multiple values from functions


<center>
**return(list(....))**
</center>

# Example - Biodiversity function

 
 * take a look at the *compute_diversity* function  - any questions about what it does?
 
# Sample data for Biodiversity Function
```{r}

# get the function
source("../R/compute_diversity.R")

# sampling categorical data
# create possible values
flower_types = c("iris","daisy","poppy","rose","dandelion","weed","violet")
# sample (with replacement)
garden = sample(flower_types, size=20, replace=T)
compute_diversity(garden)
```
 
 
 Try it
 * modify to also return the *rarest* species
 
