---
title: "simpledynamic"
output: html_document
date: '2022-04-28'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(purrr)

```

# Implementing Dynamic Models in R

Dynamic models always involves derivatives (equations that express how things change from time step to time step or place to place )


Implement population **growth** as a derivative - a model of population change

```{r derivaties, echo=TRUE}

# note that we include time here but we don't use it; we will need this later
source("../R/dexppop.R")

# see how it works
dexppop(P=20, r=0.01)


#what is this?

# notices this is the same as
dexppop(t=100,P=20, r=0.01)

# lets look at this for a range of initial populations
pops = seq(from=1, to=100)
tmp = pops %>% map(~dexppop( time=0,r=0.01, P=.x))
pchange = unlist(tmp)


pdyn = data.frame(pops, pchange)
ggplot(pdyn, aes(pops, pchange))+geom_point(col="green", size=1.5)


# why is this a straight line?
# how many new individuals are born at each population level

# try this - add a carrying capacity (dP/dt = 0 if P > carryingcapacity)
```

# Integration

What if we wanted to look at population in 20 years given an initial condition

Two options

* explicit solution to differential equation is known; e.g. you can integrate both sides of the equation! Not always possible but lets look at a case where it is possible

* must be solved by iteration; this is what we do when we can't integrate both sides 

# Explicit Solution is available


```{r explicit}
source("../R/exppop.R")

# gives population after any time given an initial population

# 20 rabbits, growth rate of 0.01 how many in 30 years
exppop(T=30, P0=20, r=0.01, K=1000)

# if we want to see how population evolves over time - generate a time series by running our model for each point in time

initialrabbits = 20
years = seq(from=1, to=100, by=2)
Ptime = years %>% map_dbl(~exppop( P0=initialrabbits, r=0.01, K=1000, T=.x))

# keep track of what times we ran
Ptime = data.frame(P=Ptime, years=years)

ggplot(Ptime, aes(years,P))+geom_point()+labs(x="years",y="Rabbit Population")

# try generating results for maximum and minimum possible r values to compare (guess at what you think)


max_r = 0.1
min_r = 0.01
K = 1000

tmp = years  %>% map_dbl(~exppop(r=max_r, P0=initialrabbits, K=K, T=.x))
Ptime$Pmaxr = tmp
tmp = years %>% map_dbl(~exppop(r=min_r, P0=initialrabbits, K=K, T=.x))
Ptime$Pminr = tmp

head(Ptime)
Ptimep = Ptime %>% gather(key="r",value="P",-years)
ggplot(Ptimep, aes(years,P, col=r))+geom_point()+labs(x="years",y="Rabbit Population")

# notice how populatin becomes unstable for high growth rates! 


```


# Solving by thinking of problem as a difference equations

Population models can be discrete (rather than continuous)

So we could implement them as difference equations and iterate


```{r iterate1}
source("../R/discrete_logistic_popK.R")
# notice how a for loop is used to iterate

# how many rabbits after 50 years given a growth of 0.1
# starting with 1 rabbit - but a carrying capcity of 500

discrete_logistic_pop

discrete_logistic_pop(P0=1, r=0.05,K=200, T=50)

# save results
discrete_result = discrete_logistic_pop(P0=1, r=0.05, K=200, T=50)

# lets also keep the parameters for use later
P0=1
r=0.05
K=200
T=50

```

# Compare discrete and analytic results

Save the results from both to compare


```{r closed}

source("../R/exppop.R")

exppop(P0=P0, r=r, K=K, T=T)
analytic_result = exppop(P0=P0, r=r, K=K, T=T)

analytic_result
discrete_result

# why are they different
# look at trajectories

growth_result = data.frame(time=seq(from=1,to=100))

growth_result$Panalytic = growth_result$time %>% map_dbl(~exppop( P0=1,r=0.05, K=200,T=.x ))

growth_result$Pdiscrete = growth_result$time %>% map_dbl(~discrete_logistic_pop( P0=1,r=0.05, K=200,T=.x ))

tmp = growth_result %>% gather(key="Ptype",value="P",-time)
ggplot(tmp, aes(time,P, col=Ptype))+geom_point()

# try running them for longer time periods to see what happens 
# change the value of r, K , P0 - see how it effects the results
```


# ODE example
```{r usingode}


source("../R/dexppop.R")

dexppop
library(deSolve)
initialrabbits = 20
years = seq(from=1, to=100, by=2)

# run the solver
Ptime = ode(y=initialrabbits, times=years, func=dexppop,parms=c(0.01))
head(Ptime)
colnames(Ptime)=c("year","P")

# notice that there are additional pieces of information year, including the method used for integration
attributes(Ptime)

# this also means you need to extract just the data frame for plotting
ggplot(as.data.frame(Ptime), aes(year,P))+geom_point()+labs(y="Population", "years")

# this also works (of course function can be by order)
Ptime=ode(initialrabbits, years, dexppop,0.03)
colnames(Ptime)=c("year","P")
ggplot(as.data.frame(Ptime), aes(year,P))+geom_point()+labs(y="Population", "years")


```
