---
title: "sobol_with_ode"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(deSolve)
library(sensitivity)
```

```{r sen}
source("../R/dpopgrowth.R")

dpopgrowth

# lets start with sobel 
library(sensitivity)

# come up with first set of sample parameters
# we will assume that we know the initial population,

Pinitial=10

# want to learn about sensitivity to growth rate (r) and carrying capacity 
# set the number of parameters
np=2000
K = rnorm(mean=200, sd=50, n=np)
r = rnorm(mean=0.05, sd=0.01, n=np)
X1 = cbind.data.frame(r=r, K=K)

# repeat to get our second set of samples
K = rnorm(mean=200, sd=50, n=np)
r = rnorm(mean=0.05, sd=0.01, n=np)
X2 = cbind.data.frame(r=r, K=K)

# fix any negative values and they are not meaningful
X1 = X1 %>% map_df(pmax, 0.0)
X2 = X2 %>% map_df(pmax, 0.0)

# create our sobel object and get sets ofparameters for running the model

sens_P = sobolSalt(model = NULL,X1, X2, nboot = 300)

# our parameter sets are
head(sens_P$X)

# lets add names 
colnames(sens_P$X) = c("r","K")

head(sens_P$X)


sens_P$X[1,]
# recall ODE needs ALL of our parameters in a single list 
# initial population and times for which we want output 
Pinitial

# gets results for 200 years (evaluating every year)
simtimes = seq(from=1, to=200)
parms = list(r=sens_P$X[1,"r"], K=sens_P$X[1,"K"])

# or
parms = list(r=as.data.frame(sens_P$X)$r[1], K=as.data.frame(sens_P$X)$K[1])

result = ode(y=Pinitial, times=simtimes, func=dpopgrowth, parms=parms)

head(result)
colnames(result)=c("time","P")
# turn it into a data frame
result = as.data.frame(result)
ggplot(result, aes(time, P))+geom_point()

# extra our metrics of interest  from this
# maximum population it gets to
maxpop = max(result$P)
maxpop

# years required to get to a threshold population (150)
# which will tell when this occurs - we will take the first one
thresh = 150
idx = which(result$P > thresh)[1]

# if it never gets there
idx = ifelse(is.na(idx), length(result$P), idx)
# turn this index into a year (might be the same if time step in 1 but just in case it isn't)
threshyear = result$time[idx]
threshyear

ggplot(result, aes(time, P))+geom_line()+geom_vline(xintercept=threshyear, col="red")+labs(y="Population",title="When do we get to 150")

# or how about threshold of 50% of carrying capacity
thresh = 0.5*sens_P$X[1,"K"]
idx = which(result$P > thresh)[1]

# if it never gets there
idx = ifelse(is.na(idx), length(result$P), idx)
# turn this index into a year (might be the same if time step in 1 but just in case it isn't)
threshyear = result$time[idx]
threshyear

ggplot(result, aes(time, P))+geom_line()+geom_vline(xintercept=threshyear, col="red")+labs(y="Population",title="When do we get to 50%\nof carrying capacity")

```
# Compute our metric for all the parameter sets

What if we want to run for all parameters

Lets create two additional functions that will help us

* a function that computes the metrics we want

* a function that runs our ode solver and computes the metrics (I call it a wrapper function as it is really just a workflow/wrapper to call ode solver and then compute metrics)
```{r sen2}
# turn computing our metrics into a function

compute_metrics = function(result, thresh) {
  maxpop = max(result$P)
idx = which(result$P > thresh)[1]
idx = ifelse(is.na(idx), length(result$P), idx)
threshyear = result$time[idx]
return(list(maxpop=maxpop, threshyear=threshyear))}

# try it on our first parameter set, and look at when it gets to 100
compute_metrics(result, 100)

# great but we need to apply the ode and this function for all of our parameters



# define a wrapper function to do everything we need - run solver and compute metrics - and send back results for each parameter

# lets make the threshold 90% of carrying capacity

p_wrapper = function(r,K, Pinitial, simtimes, odefunc, metricfunc) {
    parms = list(r=r, K=K)
    result = ode(y=Pinitial, times=simtimes, func=odefunc, parms=parms) 
    colnames(result)=c("time","P")
  # get metrics
  metrics=metricfunc(as.data.frame(result), thresh=100)
  return(metrics)
}

# test
p_wrapper(r=0.01, K=150, Pinitial=3, simtimes=seq(from=1,to=10),
         odefunc=dpopgrowth, metricfunc=compute_metrics)
```



# Next step

Run the wrapper for all parameters and look at results


```{r userwarpper}



# now use pmap as we did before

allresults = as.data.frame(sens_P$X) %>% pmap(p_wrapper, Pinitial=Pinitial, simtimes=simtimes, odefunc=dpopgrowth, metricfunc=compute_metrics)

# extract out results from pmap into a data frame
allres = allresults %>% map_dfr(`[`,c("maxpop","threshyear"))


# create boxplots
tmp = allres %>% pivot_longer(cols=everything(),names_to="metric", values_to="value")
ggplot(tmp, aes(metric, value, col=metric))+geom_boxplot()

```

# Compute the sobol indicies for each metric

```{r sen3}
# sobol can only handle one output at a time  - so we will need to do them separately

sens_P_maxpop = sensitivity::tell(sens_P,allres$maxpop)

# first-order indices (main effect without co-variance)
rownames(sens_P_maxpop$S)=c("r","K")
sens_P_maxpop$S

# total sensitivity index -note that this partitions the output variance 
rownames(sens_P_maxpop$T)=c("r","K")
sens_P_maxpop$T




# create another one for max year
sens_P_threshyear = sensitivity::tell(sens_P,allres$threshyear)
# first-order indices (main effect without co-variance)
rownames(sens_P_threshyear$S)=c("r","K")
sens_P_threshyear$S

# total sensitivity index -note that this partitions the output variance - so values sum to 1
rownames(sens_P_threshyear$T)=c("r","K")
sens_P_threshyear$T

```

